### 2015-06-08 chapters/introduction.tex
minor corrections in section encoding

There are many in-depth histories of the origin and development of writing
systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
not repeat here. However, the history of turning writing into computer readable
code is not so often told, so we decided to offer a short survey of the major
developments of such \textsc{encoding} here.\footnote{Because of the recent
history as summarized in this section, we have used mostly rather ephemeral
internet sources. When not references by traditional literature in the
bibliography, we have used \url{http://www.unicode.org/history/} and various
Wikipedia pages for the information presented here. A useful survey of the
historical development of the physical hardware of telegraphy and
telecommunication is \citet{Huurdeman2003}. Most books that discuss the
development of encoding of telegraphic communication focus of cryptography,
e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
i.e.~non-cryptographic, encoding that is related here.} This history turns out
to be intimately related to the history of telegraphic communication.

@@

Telegraphy was reinvigorated in the end of the eighteenth century through the
introduction of so-called semaphoric systems by Claude Chapelle to convey
messages over large distances. Originally, various specially designed
contraptions were used to send messages. Today, descendants of these systems
are still in use utilizing flags or flashing lights. The ``innovation'' of those
semaphoric systems was that all characters of the written language were replaced
one-to-one by visual signals. Since then, all telegraphic systems have taken
this principle, namely that any language to be transmitted first has to be
turned into some orthographic system, which subsequently is encoded for
transmission by the sender, and then turned back into orthographic
representation at the receiver side.\footnote{Sound and video-based
telecommunication of course takes a different approach by ignoring the written
version of language and directly encode sound waves or light patterns.} This of
course implies that the usefulness of any such telegraphic encoding completely
depends on the sometimes rather haphazard structure of orthographic systems.

@@

From those bisignal encodings, true \textsc{binary codes} developed with a fixed
length of signals per character. In such systems only a single separator between
signals is needed, because the separation between characters can be established
by counting until a fixed number of signals has passed.\footnote{Of course, no
explicit separator is needed when the timing of the signals is known, which is
the principle used in all modern telecommunication systems. An important modern
consideration is also how to know where to start counting when you did not catch
the start of a message, something that is known in Unicode as \textsc{self
synchronization}.} In the context of electric telegraphy, such a binary code
system was first established by Émile Baudot in 1870, using a fixed combination
of five signals for each written character.\footnote{True binary codes have a
longer history, going at least back to the Baconian cipher devised by Francis
Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
leading to all modern systems.} There are $2^5 = 32$ possible combination when
using five binary signals; an encoding today designated as 5-bit. These
codes are sufficient for all Latin letters, but of course they do not suffice
for all written symbols, including punctuation and digits. As a solution, the
Baudot code uses a so-called ``shift'' character, which signifies that from
that point onwards---until shifted back---a different encoding is used, allowing
for yet another set of 32 codes. In effect, this means that the Baudot code, and
the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an
extra bit of information, so the encoding is actually 6-bit (with $2^6
= 64$ different possible characters). For decades, this encoding was the
standard for all telegraphy and it is still in limited use today.

@@

Because 256 characters were clearly not enough for the many different characters
needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
extended the MCS to include 16 different 8-bit code pages. For example, part 5
was used for Cyrillic characters, part 6 for Arabic, and part 7 for
Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
actually an $8+4=12$-bit encoding, though with very many duplicates by design, namely 
all ASCII codes were repeated in each 8-bit code page. To be precise,
ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
characters.} This system almost immediately was understood to be insufficient and
impractical, so various initiatives to extend and reorganize the encoding
started in the 1980s. This led, for example, to various proprietary encodings
from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one
still sometimes encounters today.

>>>>

There are many in-depth histories of the origin and development of writing
systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
not repeat here. However, the history of turning writing into computer readable
code is not so often told, so we decided to offer a short survey of the major
developments of such encoding here.\footnote{Because of the recent
history as summarized in this section, we have used mostly rather ephemeral
internet sources. When not references by traditional literature in the
bibliography, we have used \url{http://www.unicode.org/history/} and various
Wikipedia pages for the information presented here. A useful survey of the
historical development of the physical hardware of telegraphy and
telecommunication is \citet{Huurdeman2003}. Most books that discuss the
development of encoding of telegraphic communication focus of cryptography,
e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
i.e.~non-cryptographic, encoding that is related here.} This history turns out
to be intimately related to the history of telegraphic communication.

@@

Telegraphy was reinvigorated in the end of the eighteenth century through the
introduction of so-called semaphoric systems by Claude Chapelle to convey
messages over large distances. Originally, various specially designed
contraptions were used to send messages. Today, descendants of these systems are
still in limited use, for example utilizing flags or flashing lights. The
``innovation'' of those semaphoric systems was that all characters of the
written language were replaced one-to-one by visual signals. Since then, all
telegraphic systems have taken this principle, namely that any language to be
transmitted first has to be turned into some orthographic system, which
subsequently is encoded for transmission by the sender, and then turned back
into orthographic representation at the receiver side.\footnote{Sound and
video-based telecommunication of course takes a different approach by ignoring
the written version of language and directly encode sound waves or light
patterns.} This of course implies that the usefulness of any such telegraphic
encoding completely depends on the sometimes rather haphazard structure of
orthographic systems.

@@

From those bisignal encodings, true \textsc{binary codes} developed with a fixed
length of signals per character. In such systems only a single separator between
signals is needed, because the separation between characters can be established
by counting until a fixed number of signals has passed.\footnote{Of course, no
explicit separator is needed at all when the timing of the signals is known, which is
the principle used in all modern telecommunication systems. An important modern
consideration is also how to know where to start counting when you did not catch
the start of a message, something that is known in Unicode as \textsc{self
synchronization}.} In the context of electric telegraphy, such a binary code
system was first established by Émile Baudot in 1870, using a fixed combination
of five signals for each written character.\footnote{True binary codes have a
longer history, going at least back to the Baconian cipher devised by Francis
Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
leading to all modern systems.} There are $2^5 = 32$ possible combination when
using five binary signals; an encoding today designated as 5-bit. These
codes are sufficient for all Latin letters, but of course they do not suffice
for all written symbols, including punctuation and digits. As a solution, the
Baudot code uses a so-called ``shift'' character, which signifies that from
that point onwards---until shifted back---a different encoding is used, allowing
for yet another set of 32 codes. In effect, this means that the Baudot code, and
the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an
extra bit of information, so the encoding is actually 6-bit (with $2^6
= 64$ different possible characters). For decades, this encoding was the
standard for all telegraphy and it is still in limited use today.

@@

Because 256 characters were clearly not enough for the many different characters
needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
extended the MCS to include 16 different 8-bit code pages. For example, part 5
was used for Cyrillic characters, part 6 for Arabic, and part 7 for
Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
actually an $8+4=12$-bit encoding, though with very many duplicates by design,
namely all ASCII codes were repeated in each 8-bit code page. To be precise,
ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
characters. However, because of overlap and not-assigned codes points the actual
number of symbols was much smaller.} This system almost immediately was
understood to be insufficient and impractical, so various initiatives to extend
and reorganize the encoding started in the 1980s. This led, for example, to
various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
(e.g.~Mac OS Roman), which one still sometimes encounters today.
